---
title: "Final Project-Maribel Viveros"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: "2023-11-1"
---
# Setup

```{r}
## DH: lubridate, ggplot2, dplyr, tibble all loaded by tidyverse
suppressPackageStartupMessages({
  suppressWarnings(library(tidyverse))   # for working with the data
  suppressWarnings(library(lubridate))   # for working with datetime data
  suppressWarnings(library(skimr))       # generate a text-based overview of the data
  suppressWarnings(library(visdat))      # generate plots visualizing data types and missingness
  suppressWarnings(library(plotly))      # generate interactive plots
  suppressWarnings(library(readxl))
  suppressWarnings(library(ggplot2))     # needed for the plot of eye metrics
  suppressWarnings(library(dplyr))
  suppressWarnings(library(tibble))
})
```


# Exploratory Eye Gaze Analysis

## Introduction


By analyzing an existing gaze data set collected from 24 participants engaged in eight distinct desktop activities this research will contribute to our understanding of visual perception. This project aims to assess first fixation metrics for recognizing desktop activities. The first fixation, represents the immediate processing of the attended stimulus and serves as a measure of attention.

The proposed research is specifically concerned with: Can the application of first fixation metrics (a common eye-tracking measure) contribute to desktop activity recognition? 

The dataset is derived from Srivastava, N., Newn, J., and Velloso, E.'s (2018) study, where the goal was to identify sedentary activities with minimal physical movements using eye movements. This study aimed to extend previous research showcasing the potential of eye tracking as a promising method for activity recognition. The study highlighted the ongoing challenge in the research community to achieve finer-grained activity recognition across a wide range of desktop-based activities using more precise eye-measures. Srivstava and colleagues (2018) hoped their findings would support the “design of computing systems that can proactively monitor daily activities and can either assist users with their daily tasks or encourage them to follow a healthy lifestyle.” In contrast to previous work centered on using eye movements for traditional human activity recognition, their study was oriented toward devising a methodology for distinguishing desktop activities. They constructed a classifier that combined 26 low-level gaze features from existing literature with 24 novel mid-level gaze features. The researchers applied their approach to a dataset encompassing 24 participants engaged in eight desktop computing activities. The results supported the integration of the Srivastava and colleagues’ proposed mid-level gaze features, revealing enhanced overall performance in activity recognition. The data aimed to reveal the relationship and uncover the connection between eye movements and specific desktop-based activities. The values crystallized in the data is the recognition of human activities through eye movements. These values are related to desktop activity recognition and context-aware systems and consist of high, low and mid-level gaze features and timestamps in the dataset.  
  This understanding of human activities is valuable in several domains, including surveillance, healthcare, and education. The data contributes to the development of context-aware interactive systems. By understanding the context of users' activities, systems can provide more personalized and helpful support. Ultimately, their study's objective is to simplify activity recognition by leveraging intuitive knowledge of eye movements and patterns.


The current research inquiry revolves around the question, to what extent can the incorporation of first fixation metrics enhance the recognition of desktop activities? I aim to delve into the specific eye behaviors, initial gaze, to see how it supports activity recognition. Prior studies have shown that first fixation duration can serve as a measure of visual information acquisition (Holmqvist, et al., 2018). First fixation, or initial gaze, is the amount of time an individual spends fixating their gaze on a specific area of interest during an eye-tracking experiment. Given that prior researchers have already identified eye movements that are operationalized into mid-level, low-level, and high-level categories suitable for classification, my study seeks to ascertain which features are most appropriate for this purpose.

This isn’t very precise, but that’s okay: Part of the goal of this EDA project is to clarify eye-metrics that contribute to understanding of visual perception.

```{r}
## DH: Don't use absolute paths
file_path <- file.path('data', 'desktopactivities all.xlsx')
```

##  Methods

The study will be using eye-data on students as they preformed different desktop activities, available on Kaggle. The dataset comprises raw gaze coordinates (x-y) and timestamp data collected from 24 participants who were engaged in eight distinct desktop activities: Read, Browse, Play, Search, Watch, Write, Debug, and Interpret.

The dataset is available for download  at the URL: <https://www.kaggle.com/datasets/namratasri01/eye-movement-data-set-for-desktop-activities>. There are 192 individual files, so I need to do some data wrangling and combine them into one dataframe for analysis.

<!-- DH: Okay, so what's up with the Excel spreadsheet in the `data` folder? -->

<!-- DH: You're jumping back and forth a bit between the purposes of the data collection and material details about retrieving the data from Kaggle. Give us the high-level view of the purpose of the dataset in the introduction, and here focus on downloading the data. -->

The original objective  by Srivastava et al. (2018) was to introduce a level of abstraction between low- and high-level gaze features, known as "mid-level gaze features." These mid-level gaze features don't necessitate a understanding of interface design but are based on intuitive knowledge about the types of eye movements associated with different activities. Intuitively, anyone can deduce the nature of a user's activity by examining gaze patterns. 

In this context, the data collected by Srivastava et al. (2018) was used to capture real-time eye movement data as participants undertook various activities through the Tobii Pro X2-30 eye tracker. The emphasis was on utilizing  eye movement data for identifying sedentary activities.


The dataset includes the x and y coordinates and timestamps of students' eye positions as they perform assigned tasks. To enhance the predictive power, I will convert the conditions into numeric characters and merge the datasets of individual participants. This integration will provide more robust predictions. My objective is to assess the first time to fixation, and the duration of the first fixation to identify which of these factors is most effective for desktop classification. 


```{r}
sheet_names <- excel_sheets(file_path) # Get the sheet names
data_frames <- list()# Initialize an empty list to store the data frames

## DH: This approach violates immutability, and in R the preference is to use functions like `map()` rather than loops
# Read each sheet and store the data frames in the list
for (sheet in sheet_names) {
  data_frames[[sheet]] <- read_excel(file_path, sheet = sheet)
}


combined_data <- do.call(rbind, data_frames) # Combine all data frames into one big data frame

print(combined_data) # Print the combined data

## DH: Since we're never using `data_frames` again, and it's kind of big, recommend cleaning it up
rm(data_frames)
```

# Results

For the EDA project I will use  base R functions to look at dimensions of the dataframe and column (variable) types outlined by Peng and Matsui (2016). The analysis covers the all the items from the checklist. The following results  below were found. <!-- DH: You don't use base R functions below -->

```{r}
skim(combined_data) #3. Check the packaging ~How many rows and columns?    Are there any types that might indicate parsing problems? NO
```

## Dataset

Specifically, the dataset contains information on 24 unique participants, 3 unique trials, and 8 unique activities.  <!-- DH: Are you getting these numbers from metadata on Kaggle (or some other exterior source), or from the dataset itself?  Could be a good way to check your Ns! --> The data includes rows with unique x, y, and timestamp values, with each row identified by the raw_row_number. The variables 'participant', 'set', and 'activity' serve as identifiers.  Notably, the participant, set, and activity variables do not contain any missing values. Within the dataset, the variables x and y represent the spatial coordinates, while the timestamp indicates the specific time point when the gaze was at the corresponding x, y coordinate. The timestamp is instrumental in representing durations or the "time spent" on a particular activity. It's worth noting that the variables representing activity and set also do not contain any missing values. In total the data set consists of 1,505,813 rows (observations); 6 columns (variables). The 3 variables are handled as characters, 3 as numeric. 
  For the motivating question, the data set is appropriate as the eye gaze coordinates and timestamp is 100% complete. Also, the desktop activity data is 100% complete. Some potential limitations are the complicated nature of eye-metric data in x, y and timestamp analysis, especially given the lack of dimensions of the desktop used so lets presume a 1080 resolution. Resolution is important to calculate saccades,and important in fixation duration for future studies. 
  To examine the data set for missing values I can't use 'vis_miss(combined_data)' because of large data which causes a error so 'sample_n' to draw a subset was used. Arguments in vis_miss() are useful for picking up patterns in missing values. As seen below there is no missing data.
<!-- DH: Recommend breaking this giant wall of text into multiple paragraphs.  You can also intersperse code blocks: metadata says there should be 24 unique participants, and here's how that compares to `count(combined_data, participant)`. -->

```{r}
set.seed(123)
dataf_smol = sample_n(combined_data, 1000) #sample to big so we'll draw a subset #but no missing data 

vis_miss(dataf_smol)
```


```{r}
## DH: Why take another subsample? 
combined_smol = sample_n(combined_data, 1000) #subset of combined data
```

## A Critical Examination

By examining the mean duration of the first timestamp for each activity, we can discern patterns related to the amount of time spent on each specific activity. This can help identify which activities require more prolonged or shorter periods of visual engagement. To answer the following question, Can we identify which activities require more prolonged or shorter periods of visual engagement? 


```{r}
combined_data %>%
  filter(!is.na(activity)) %>%
  arrange(timestamp) %>%
  group_by(activity) %>%
  summarise(mean_duration = mean(timestamp - first(timestamp)), #This calculates the mean duration for each group (activity) using the difference between the timestamp and the first timestamp value within each group. It provides the average duration of time spent on the initial instance of each activity.
            min_x = first(x), 
            min_y = first(y))
```

<!-- DH: So what's your answer? -->

# Data Summary-Check your n's

<!-- DH: This was step 4, "Look at the top and bottom of your data" -->

With  over 1 million rows, the dataframe is too large to print in a readable way. Instead  the base R function View() in an interactive session was used. An Excel-like spreadsheet presentation View() can cause significant problems if you use it with a large dataframe on a slower machine, so I used a pipe. Overall, this code block is designed to provide an interactive way to examine the top and bottom rows of the combined_data and combined_smol data frames, allowing for easy inspection and understanding of the big data structure and contents.

```{r}
if (interactive()) {
    combined_data |> 
        head() |> 
        View()
    
    combined_data |> 
        tail() |> 
        View()
    
    View(combined_smol) #4 Check Top and Bottom
}
```

<!-- DH: Anything notable here? -->

The skimr to check data quality was applied and examined the minimum and maximum values. All of the ranges make sense for what I expect the variable to be given each of the activities were not to exceed 5-6 minutes.

<!-- DH: You already did this, in the "Results" section? -->

```{r}
skim(combined_data)
```


Since there is not another duplicate set or typical data set for desktop activities and eye-metric times, I cannot check against n's, as the total expected are in the dataframe. So here I am checking the sample to see if timescale for one activity matches with the dataset. In this example, by downloading the dataset from the kaggle website, you can see the first, 'x' and 'y' coordinates in 'write' activity for participant no.24, assigned in set 'B' match with the sample below. 

```{r}
## DH: What is this? What are we supposed to do with it? 
# Given table
data <- tibble::tribble(
  ~participant, ~set, ~activity, ~x, ~y, ~timestamp, #plot of head
  "P24", "B", "WRITE", 930, 555, 0,
  "P24", "B", "WRITE", 629, 426, 33,
  "P24", "B", "WRITE", 224, 332, 71,
  "P24", "B", "WRITE", 199, 334, 101,
  "P24", "B", "WRITE", 214, 342, 134,
  "P24", "B", "WRITE", 224, 324, 16
)
```

# Validation

In the original methods section, the authors detail that all of the activities last about 5-6 minutes. Therefore the timestamps for the coordinates for each activity for all participants should not exceed this time, and indeed the max duration for one activity in going through all the values is ~6 minutes (399184 ms). A web search leads us to the website for the article where the data set is used: <https://www.researchgate.net/publication/329955224_Combining_Low_and_Mid-Level_Gaze_Features_for_Desktop_Activity_Recognition> and published.


```{r}
combined_data %>%
  filter(!is.na(activity)) %>%
  arrange(timestamp) %>%
  group_by(activity, participant) %>%
  summarise(
    max_duration = max(timestamp - first(timestamp)),
    min_x = first(x),
    min_y = first(y)
  ) %>%
  arrange(desc(max_duration))#By adding arrange(desc(max_duration)) at the end of the pipeline, the data will be sorted in descending order based on the max_duration column, showing the highest amounts of max_duration first.

## DH: Good, though this leaves out the step of comparing these values to the 399184 ms limit.  Compare: 
combined_data %>%
  filter(!is.na(activity)) %>%
  arrange(timestamp) %>%
  mutate(too_long = timestamp > 399184) %>% 
  count(too_long)
```

# Plot 

The initial plot illustrates the variability in the average time participants spend observing a stimulus based on different activities. To further investigate, the second graph focuses on assessing the average duration of the first fixation-first glance, which represents the time difference between the first and start of timestamps. This analysis aims to reveal more noticeable distinctions in fixation times across various activities and participants to find if there are noticible differences. 
  Finally, in the box plot graph, the dots show the individual participant first fixation duration average values for each activity.The code calculates the first fixation duration by determining the time difference between the timestamp of the first fixation and the start time (0). The resulting plot then provides the average duration of the first fixation for each group defined by 'activity' and 'participant'. This information is derived from the computation of the avg_first_fixation_duration. 

```{r}
## DH: First, geom_bar() with `stat = 'identity'` is just geom_col().  Second, using columns here results in heavy overplotting, which you can see by setting `color = 'black'`: there's one column for each activity-participant combination.  Without that you just see the column for the highest average.  What you want is geom_point(), showing the average for each participant in each activity. 
combined_data %>%
  filter(!is.na(activity)) %>%
  arrange(timestamp) %>%
  group_by(activity, participant) %>%
  summarise(
    avg_duration = mean(timestamp - first(timestamp)),
    min_x = first(x),
    min_y = first(y)
  ) %>%
  ggplot(aes(x = activity, y = avg_duration, fill = activity)) +
  # geom_col(color = 'black') +
  geom_point() +
  labs(title = "Average Duration Time Across Activities for Each Participant",
       x = "Activity",
       y = "Average Duration (ms)",
       fill = "Activity") +
  theme_minimal()

```
```{r}
## DH: Same issue here.  Here's a variant where we're connecting the averages for each participant. Looks to me like 2 or 3 participants tend to have especially long first fixation across activities. 
combined_data %>%
  filter(!is.na(activity)) %>%
  arrange(participant, activity, timestamp) %>%
  group_by(activity, participant) %>%
  summarise(
    avg_duration = mean(timestamp - first(timestamp)),
    min_x = first(x),
    min_y = first(y),
    avg_first_fixation_duration = mean(c(0, diff(timestamp)))  # Calculate average duration of the first fixation
  ) %>%
  ggplot(aes(x = activity, y = avg_first_fixation_duration, 
             color = participant)) +
  # geom_col() +
  geom_line(aes(group = participant), alpha = .25) +
  geom_point() +
  scale_color_discrete(guide = 'none') +
  labs(title = "Average Duration of First Fixation Across Activities for Each Participant",
       x = "Activity",
       y = "Average Duration of First Fixation (ms)",
       fill = "Activity") +
  theme_minimal()
```
# A plot that tells us location for each activity.

To translate the research question: Do all participants stare in relatively the same positions across the activities? The easy solution is to estimate location by plotting the x, or y coordinates since eye-movements are symmetrical, for each participant across the average first fixation time stamp for each activity. Only the row with the minimum timestamp (i.e., the first fixation) is provided below by each activity. The preliminary results show, the first fixation across desktop activites is different but most participants look in the same area dependent on the activity. The plot below summarizes data related to participants' activities, then creates a boxplot, showing the distribution of the first x-coordinate across activities for each participant and the average duration of the first fixation. The plot is faceted by activity for better comparison.


```{r}
## DH: Having trouble reading this with scatterplots overlaying variable-width histograms.  Seems like this would work better as two separate plots. 
combined_data %>%
  filter(!is.na(activity)) %>%
  arrange(participant, activity, timestamp) %>%
  group_by(activity, participant) %>%
  summarise(
    avg_duration = mean(timestamp - first(timestamp)),
    first_x = first(x),  # Take the first x-coordinate
    min_y = first(y),
    avg_first_fixation_duration = mean(c(0, diff(timestamp)))  # Calculate average duration of the first fixation
  ) %>%
  ## DH: long line
  ggplot(aes(x = avg_first_fixation_duration, 
             y = first_x, 
             fill = activity)) +
  geom_boxplot() +
  ## DH: long line; remove unnecessary comment
  geom_point(position = position_jitter(height = 1), 
             size = .02, 
             alpha = 1) + 
  labs(title = "Distribution of First Fixation Durattion and X-Coordinate Across Activities for Each Participant",
       x = "Average First Fixation Duration (ms)",
       y = "first x coordinate",
       fill = "Activity") +
  theme_minimal()+
  ## DH: long line, spacing; why the x suffix? 
  scale_x_continuous(labels = scales::number_format(scale = 2, 
                                                    suffix = "x")) +
  ## DH: Consider using scales = 'free_x'
  facet_wrap(~ activity)
```

```{r}
combined_data %>%
  filter(!is.na(activity)) %>%
  arrange(participant, activity, timestamp) %>%
  group_by(activity, participant) %>%
  summarise(
    first_x = first(x),  # Take the first x-coordinate
    first_y = first(y)   # Take the first y-coordinate
  ) %>%
  ggplot(aes(x = first_x, y = first_y, color = activity)) +
  geom_point(position = position_jitter(height = 1), size = 2, alpha = 0.7) +
  labs(
    title = "Distribution of First X and Y Coordinates Across Activities for Each Participant",
    x = "First x coordinate",
    y = "First y coordinate",
    color = "Activity"
  ) +
  theme_minimal()+
  facet_wrap(~ activity)
```

<!-- DH: Anything notable here? -->

```{r}
combined_data %>%
  filter(!is.na(activity)) %>%
  arrange(participant, activity, timestamp) %>%
  group_by(activity, participant) %>%
  summarise(
    avg_duration = mean(timestamp - first(timestamp)),
    first_x = first(x),  # Take the first x-coordinate
    min_y = first(y),
    avg_first_fixation_duration = mean(c(0, diff(timestamp)))  # Calculate average duration of the first fixation
  ) %>%
  ggplot(aes(x = avg_first_fixation_duration, y = first_x, color = activity)) +
  geom_line(size = 1) +  # Line graph for first x coordinates
  geom_point(position = position_jitter(height = 1), size = .02, alpha = 1) +  # Add individual points with jitter
  labs(
    title = "Distribution of First Fixation Duration and X-Coordinate Across Activities for Each Participant",
    x = "Average First Fixation Duration (ms)",
    y = "First x Coordinate",
    color = "Activity"
  ) +
  theme_minimal() +
  scale_x_continuous(labels = scales::number_format(scale = 2, suffix = "x")) +
  ## DH: Consider scales = 'free_x' here? 
  facet_wrap(~ activity)

```

<!-- DH: Anything notable here? -->



## Discussion

The resulting graph demonstrates the relationships between the 'activity' and 'participant' variables and average of first fixation duration. It plots the participants average first fixation time on the x-axis and utilizes facets to categorize the data points based on different activities. This facet arrangement allows for a clear comparison of the gaze patterns and x coordinates among the participants for each specific activity. Through this visual representation, we can observe distinct variations in the participants' gaze patterns across different activities and first time fixation on which area. For each activity we observe the average time spent on the first attended item and where on the item participants looked. The clear differentiation in the plotted data points suggests potential differences in how participants engage with various desktop activities, providing valuable insights into their cognitive processes and task engagement strategies during the experiment or study.

## Observations

Due to the substantial size of the data set per individual, particularly with eye-metrics generating extensive data, I anticipated challenges in the analysis. Determining which instance of "first time to fixation" to utilize proved difficult. Typically, eye tracking studies rely on the averages of first time to fixation for analysis so that is what I included. The complexity of the data aligned with my expectations and it was a challenging project. However, some notable differences in location of first fixation and time across activity were observed so the data fit the research question and I was able to make some conclusions.  In shaping the trajectory of future research, the focus should extend beyond the averages of first time to fixation.  The box plot revealed outliers beyond the mean, upper, and lower quartiles for each activity, indicating a need for focused investigation in individual differences.  Therefore, future studies should prioritize examining and understanding these individual differences, guiding the research community towards a more nuanced and impactful understanding of eye-metrics. My current research focus is understanding individual differences in eye-patterns and behaviors.



## Citations

Holmqvist, K., Nyström, M., Andersson, R., Dewhurst, R., Halszka, J., & van de Weijer, J. (2011). Eye Tracking : A Comprehensive Guide to Methods and Measures. Oxford University Press. http://ukcatalogue.oup.com/product/9780199697083.do

Srivastava, N., Newn, J., & Velloso, E. (2018). Combining Low and Mid-Level Gaze Features for Desktop Activity Recognition. Proceedings of the ACM on Interactive, Mobile, Wearable and Ubiquitous Technologies, 2(4), 189.

